Real Time Variational Density Forecasting

### 2017-02-22 Confirmation/confirmation_V2.tex
final tweaks before draft submission

There are many types of MCMC algorithms, with arguably the simplest and most commonly used one being the Gibbs sampler. The Gibbs sampler algorithm iteratively samples the components of the $k-$dimensional parameter vector $\boldsymbol{\theta}$ via each of the so-called full conditional distributions as follows,
\begin{align}
&p(\theta_1 | \theta_2, \dots, \theta_k, y_{1:T}) \nonumber \\
&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_k, y_{1:T}) \nonumber \\
&\vdots \nonumber \\
&p(\theta_k | \theta_1, \dots, \theta_{k-1}, y_{1:T}). \nonumber
\end{align}
Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\boldsymbol{\theta} | y_{1:T})$ and can be used for forecasting and inference about the unknown parameters. The computation time for each iteration and the overall number of iterations required to accruately summarise the posterior distribution is problem specific and typically increases with the number of parameters in the model. On occasion, the full conditional distribution can not be recognised and a Metropolis-Hastings-within-Gibbs step can be used, see \citet{Gilks1995} for details.

@@

Figure (\ref{fig:norminvg}) compares the simulated posterior densities from Section (\ref{sec:MCMC}), (red) with the approximation from Mean Field Variational Bayes (blue) and the analytically available true posterior marginal densities (black). For both parameters all three densities are indistinguishable.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{norminvg}
\caption{Marginal posterior densities for the model described in Section (\ref{sec:MCMC}). The true density is in black, the Monte Carlo simulation is red and the MFVB approximation is blue. Both MCMC and MFVB fit this density extremely well.}
\label{fig:norminvg}
\end{figure}

@@

This idea is extended by the vine copula, a technique to factorise a high-dimensional copula into a set of bivariate copulas, see \citet{Joe2014} and references within. A vine copula is represented by a series of trees, where the top-most tree is a graph that contains each variable as a node. The edges in each tree then form the nodes in the tree below, which are connect by an edge only if the associated edges in the previous tree connected to the same node. This sequence is continued or a $k$ dimensional $\boldsymbol{\theta}$ until the $(k-1)'th$ tree is reached that contains only one edge. Each of the $k(k-1)/2$ edges represents a bivariate copula between the two unique parameters in the connected nodes, conditioned on every parameter appearing in both nodes. Refer to Figure (\ref{fig:vinecop}) for an example.

\iffalse
\begin{figure}[h]
\centering
\tikz{ %
    \node[latent] (theta1) {$\theta_1$} ; %
    \node[latent, right = of theta1] (theta2) {$\theta_2}$} ; %
    \node[latent, above = of theta2] (theta3) {$\theta_3$} ; %
    \node[latent, below = of theta2] (theta4) {$\theta_4$} ; %
    \edge{theta1} {theta2} ; %
    \edge{theta2} {theta3} ; %
    \edge{theta2} {theta4} ; %

    \node[latent, below = of theta4] (theta2theta3) {$\theta_2 \theta_3$} ; %
    \node[latent, left = of theta2theta3] (theta1theta2) {$\theta_1 \theta_2$} ; %
    \node[latent, right = of theta2theta3] (theta2theta4) {$\theta_2 \theta_4$} ; %
    \edge{theta2theta3} {theta1theta2} ; %
    \edge{theta2theta3} {theta2theta4} ; %

    \node[latent, below = of theta1theta2] (t1t3t2) {$\theta_1 \theta_3 | \theta_2$} ; %
    \node[latent, right = of t1t3t2] (t3t4t2) {$\theta_3 \theta_4 | \theta_2$} ; %
    \edge{t1t3t2} {t3t4t2} ;%
}
\caption{An example of a Vine Copula over four variables. The edges in the top tree contains unconditional bivariate copulas between the parameters on the connected nodes. The second tree contains the bivariate copulas for $\theta_1 \theta_3 | \theta_2$ and for $\theta_3 \theta_4 | \theta_2$, the conditioning on $\theta_2$ caused by its appearance in each connected node. The edge in the final tree contains the bivariate copula for $\theta_1, \theta_4 | \theta_2, \theta_3$, as both $\theta_2$ and $\theta_3$ appear in the connected nodes.
\label{fig:vinecop}
\end{figure}
\fi

@@

Consider the second order auto-regressive time series model, denoted by AR(2) and described by
\begin{equation}
\label{AR2}
y_t = \phi_1 y_{t-1} + \phi_2 y_{t-2} + \epsilon_t
\end{equation}
where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$ for $t = 3, \dots, T$. The first two observations, $y_1$ and $y_2$ are distributed according to 
\begin{equation}
\label{AR2initial}
y_1, y_2, \sim \mathcal{N}(\boldsymbol{0}, \Sigma)
\end{equation}
where 
\begin{equation}
\Sigma = \left[ \begin{array}{cc} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{array} \right],\
\end{equation}
and $\gamma(k)$ is the $k-th$ autocorrelation for an AR(2) model.
The likelihood of parameters $\boldsymbol{\theta} = (\phi_1, \phi_2, \sigma^2)'$, that is, the density of the observations $y_{1:T}$, given $\boldsymbol{\theta}$, viewed as a function of $\boldsymbol{\theta}$, is given by
\begin{align}
\label{likelihood}
L(\theta | y_{1:T}) &= p(y_1, y_2 | \theta) \prod_{t=3}^{T}p(y_t | y_{1:t-1}, \theta) \nonumber \\
&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} \left( (y_1, y_2)' \Sigma^{-1} (y_1, y_2) + \frac{1}{\sigma^2} \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}.
\end{align}

@@

Finally, Variational Bayes only becomes a viable method for forecast densities if the statistical error associated with using an approximate distribution is less than the statistical error associated with forecasting based on the MCMC generated posterior distribution that is not conditioned on the most recent data.

>>>>

There are many types of MCMC algorithms, with arguably the simplest and most commonly used one being the Gibbs sampler. The Gibbs sampler algorithm iteratively samples the components of the $k-$dimensional parameter vector $\boldsymbol{\theta}$ via each of the so-called full conditional distributions as follows,
\begin{align}
&p(\theta_1 | \theta_2, \dots, \theta_k, y_{1:T}) \nonumber \\
&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_k, y_{1:T}) \nonumber \\
&\vdots \nonumber \\
&p(\theta_k | \theta_1, \dots, \theta_{k-1}, y_{1:T}). \nonumber
\end{align}
Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\boldsymbol{\theta} | y_{1:T})$. Samples taken before the MCMC converges ot the posterior must be discarded, and the remaining samples may have strong dependence between consecutive draws of the same parameter due to the Markov nature of the algorithm. The computation time for each iteration and the overall number of iterations required to accruately summarise the posterior distribution is problem specific and typically increases with the number of parameters in the model. On occasion, the full conditional distribution can not be recognised and a Metropolis-Hastings-within-Gibbs step can be used, see \citet{Gilks1995} for details.

@@

Data was simulated according to the Gaussian model introduced in Section (\ref{sec:MCMC}) with $T = 100, \mu = 2 and \sigma^2 = 1$. The hyperparameters were set to $\gamma = 0, \tau = 1, \alpha = 1$ and $\beta = 1$. Figure (\ref{fig:norminvg}) compares the analytically available true posterior marginal densities for both $\mu$ and $\sigma^2$ (black) with the densities generated by the Monte Carlo sampler (red) and the approximation from Mean Field Variational Bayes (blue). For both parameters all three densities are indistinguishable.

\begin{figure}[h]
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{norminvg}
\caption{Marginal posterior densities for the model described in Section (\ref{sec:MCMC}). The true density is in black, the Monte Carlo sampler is red and the MFVB approximation is blue. Both MCMC and MFVB fit this density extremely well.}
\label{fig:norminvg}
\end{figure}

@@

This idea is extended by the vine copula, a technique to factorise a high-dimensional copula into a set of bivariate copulas, see \citet{Joe2014} and references within. A vine copula is represented by a series of trees, where the top-most tree is a graph that contains each variable as a node. The edges in each tree then form the nodes in the tree below, which are connected by an edge only if the associated edges in the previous tree shared the same node. This sequence is continued or a $k$ dimensional $\boldsymbol{\theta}$ until the $(k-1)'th$ tree is reached that contains only one edge. Each of the $k(k-1)/2$ edges represents a bivariate copula between the two unique parameters in the connected nodes, conditioned on every parameter appearing in both nodes. Refer to Figure (\ref{fig:vinecop}) for an example.

\begin{figure}[h]
\centering
\includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{vines.png}
\caption{An example of a Vine Copula over four variables. The edges in the top tree represent unconditional bivariate copulas between the parameters on the connected nodes. The second tree edges represent the bivariate copulas for $\theta_1 \theta_3 | \theta_2$ and for $\theta_3 \theta_4 | \theta_2$, the conditioning on $\theta_2$ caused by its appearance in each connected node. The edge in the final tree represents the bivariate copula for $\theta_1, \theta_4 | \theta_2, \theta_3$, as both $\theta_2$ and $\theta_3$ appear in the connected nodes.}
\label{fig:vinecop}
\end{figure}

@@

Consider the second order auto-regressive time series model, denoted by AR(2) and described by
\begin{equation}
\label{AR2}
y_t = \phi_1 y_{t-1} + \phi_2 y_{t-2} + \epsilon_t
\end{equation}
where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$ for $t = 3, \dots, T$. The first two observations, $y_1$ and $y_2$ are distributed according to 
\begin{equation}
\label{AR2initial}
y_1, y_2, \sim \mathcal{N}(\boldsymbol{0}, \Sigma)
\end{equation}
where 
\begin{equation}
\Sigma = \left[ \begin{array}{cc} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{array} \right],\
\end{equation}
and $\gamma(k)$ is the $k-th$ autocorrelation for an AR(2) model.
The likelihood of parameters $\boldsymbol{\theta} = (\phi_1, \phi_2, \sigma^2)'$, that is, the density of the observations $y_{1:T}$, given $\boldsymbol{\theta}$, viewed as a function of $\boldsymbol{\theta}$, is given by
\begin{align}
\label{likelihood}
L(\theta | y_{1:T}) &= p(y_1, y_2 | \theta) \prod_{t=3}^{T}p(y_t | y_{1:t-1}, \theta) \nonumber \\
&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} (y_1, y_2) \Sigma^{-1} (y_1, y_2)' \right\} \nonumber \\
&\times \exp \left\{ \frac{-1}{2\sigma^2} \left( \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}.
\end{align}

@@

Finally, Variational Bayes only becomes a viable method for density forecasts if the statistical error associated with using an approximate distribution is less than the statistical error associated with forecasting based on an MCMC generated posterior distribution that is not conditioned on the most recent data. The investigation of this trade-off is of central importance to the thesis.
